---
title: "Building a Win Probability Model with R and hockeyR"
output: html_notebook
---

# Building a Win Probability Model with R and hockeyR

By :: Trevor Rowland

Based on [anpatton](<https://github.com/anpatton>)'s 
guide on [How to Make a Win Probability Model](<https://github.com/anpatton/basic-nba-tutorials/blob/main/win_probability/make_win_probability_model.md>) 
and using the [hockeyR](<https://github.com/danmorse314/hockeyR>) package.

## Step 1: Installing Packages

We will be using the tidyverse, arrow and knitr packages 
along with the sportyR and hockeyR packages for the 
necessary hockey data.

```{r}
# Install R Packages
install.packages("tidyverse")
install.packages("arrow")
install.packages("knitr")
install.packages("sportyR")
```

```{r}
# Install hockeyR from github
install.packages("devtools")
library("devtools")
devtools::install_github("danmorse314/hockeyR")
```

```{r}
# load the libraries
library("tidyverse")
library("arrow")
library("knitr")
library("hockeyR")
library("sportyR")
```

## Step 2: Generate the Dataset

### 2.a. Access the Dataset

By calling the `load_pbp(season)` method, we can 
access play-by-play data for 

```{r}
# Load Play-by-Play data into a DF
pbp_hist <- load_pbp(season = 2023)
pbp_curr <- load_pbp('2024-25')

write.csv(pbp_hist, "/Users/dB/Desktop/fall_24/DS-4210/final-proj/pbp_hist_data_2011-2023.csv", row.names=FALSE)
write.csv(pbp_curr, "/Users/dB/Desktop/fall_24/DS-4210/final-proj/pbp_curr_data.csv", row.names=FALSE)
```

The play-by-play data loaded season data from 
2011 to 2023, but was unable to pull any data 
from the most recent season. This will be resolved, 
but for now let's create the model.

### 2.b. Examine the Dataset

First we need to examine the dataset and change 
column names/clean the data when necessary.

```{r}
# Check Column Names
colnames(pbp_hist)
head(pbp_hist)

# Get all unique games
games <- unique(pbp_hist$event_id)
```

### 2.c. Testing Sample Functions

The `hockeyR` git repo provides us with some 
functions to try. Here is one that finds player 
goals above expectation:

```{r}
pbp_hist %>%
  filter(event_type %in% c("SHOT","MISSED_SHOT","GOAL")) %>%
  filter(season_type == "R" & period_type != "SHOOTOUT") %>%
  group_by(player = event_player_1_name, id = event_player_1_id, season) %>%
  summarize(
    team = last(event_team_abbr),
    goals = sum(event_type == "GOAL"),
    xg = round(sum(xg, na.rm = TRUE),1),
    gax = goals - xg,
    .groups = "drop"
  ) %>%
  arrange(-xg) %>%
  slice(1:10)
```

## Step 3: 